Nondeterministic Bigraphical Reactive Systems for Markov Decision Processes\thanks{Supported by an EPSRC Vacation Scholarship.}

### 2018-08-18 paper.tex
The paper is in a submittable state for the first time!

The full BigraphER code for this example is in Listing~\ref{example_listing}.
Reaction rule probabilities are represented as floating-point numbers inside the
arrows (e.g. \texttt{-[0.7]->}), each action encompasses its reaction rules with
\texttt{\textbf{action} actionName} and \texttt{\textbf{end}}. Rewards can be
added to actions simply by inserting an integer enclosed in square brackets
after the action name (e.g. \texttt{\textbf{action} b[1]}). Lastly, predicate
rewards have the same format, but are listed on the \texttt{\textbf{preds}} line
of the \texttt{\textbf{begin nbrs}}---\texttt{\textbf{end}} block (e.g.
\texttt{\textbf{preds} = \{heads[3]\};}).

@@

\subsection{Collecting Objects in a Grid}

@@

Now that we have two models defined in BigraphER, we will briefly show how they
can be exported and used for quantitative analysis with PRISM. More information
about the syntax and capabilities of PRISM properties can be found in the PRISM
manual\footnote{\url{http://www.prismmodelchecker.org/manual/PropertySpecification/Introduction}}.

@@

TODO (possibly): state transition diagrams and statistics about the number of
states, etc.

\begin{itemize}
\item LIMITATIONS
  \begin{itemize}
  \item no easy way to change probabilities depending on the state
  \item each room only has a single entrance/exit (could be fixed via
    bigraphs with sharing)
  \item can't implement everything you can think of: no way to check for set
    membership, limited support for representing numbers (adding/subtracting $n$
    has to be a separate reaction rule for each $n$)
  \item PRISM doesn't support importing transition rewards, positive and
    negative rewards at the same time, importing SMGs.
  \end{itemize}
\item EXTENSIONS
  \begin{itemize}
  \item discovery of spaces independent of room structure (no specific ideas
    yet, but could be very nice)
  \item there's no good reason why rewards are assigned to actions rather than
    reaction rules
  \item multiple agents
  \item uncertainty in actions (like in Artificial Intelligence: A Modern
    Approach)
  \end{itemize}
\item ADVANTAGES
  \begin{itemize}
  \item less code than with PRISM
  \item more visual and intuitive, represents the modelled situation in a more
    straightforward way
  \item rewards and transition rules can be generated with a higher degree of
    automation (considering a specific property of a state, rather than listing
    all states that satisfy that property)
  \item more flexible in changing the size of the grid
  \item can be easily extended to more than four directions
  \end{itemize}
\end{itemize}

\section{Conclusion}

Food for thought: what models have efficient (i.e. polynomial in size)
representation as BRSs? Can we define complexity classes (easy for everyone,
easy for bigraphs but not for everyone, hard for everyone)?

>>>>

The full BigraphER code for this example is in Listing~\ref{example_listing}.
Reaction rule probabilities are represented as floating-point numbers inside the
arrows (e.g. \texttt{-[0.7]->}), each action encompasses its reaction rules with
\texttt{\textbf{action} actionName} and \texttt{\textbf{end}}. Rewards can be
added to actions simply by inserting an integer enclosed in square brackets
after the action name (e.g. \texttt{\textbf{action} b[1]}). Lastly, predicate
rewards have the same format, but are listed on the \texttt{\textbf{preds}} line
of the \texttt{\textbf{begin nbrs}}---\texttt{\textbf{end}} block (e.g.
\texttt{\textbf{preds} = \{heads[3]\}}).

@@

\subsection{Collecting Objects in a Grid} \label{first_agent_section}

@@

Now that we have two models defined in BigraphER, we will briefly show how they
can be exported and used for quantitative analysis with PRISM. Suppose a model
is in a file called \texttt{agent.big}. Then, in order to export labels, the
transition matrix, and state rewards, one would call
\[ \texttt{bigrapher full -l
    agent.csl -p agent.tra -r agent.rews agent.big}. \]
The \texttt{agent.csl} file can then be augmented with the desired
\emph{properties} to check or compute, and run with PRISM with the command
\begin{multline*}
\texttt{prism -importtrans agent.tra -importstaterewards agent.rews
  \textbackslash} \\ \texttt{agent.csl}.
\end{multline*}
More information about the syntax and capabilities of PRISM properties can be
found in the PRISM
manual\footnote{\url{http://www.prismmodelchecker.org/manual/PropertySpecification/Introduction}}.

@@

%TODO (possibly): state transition diagrams and statistics about the number of
%states, etc.

While using bigraphs to model movement did allow us to achieve some results
quite easily as well as to explore ideas not easily expressed otherwise, we have
to acknowledge some limitation that we have encountered throughout the process.

First, the probability assigned to a reaction rule stays constant. The only way
it can be changed is by manipulating the number of rules applicable to a given
bigraph.

Second, there are some quite elementary concepts that cannot be
expressed in terms of bigraphs and reaction rules without significant
limitations or hindrances (or at all). For example, suppose we represent two
numbers $m, n \in \mathbb{N}$ in our bigraph and wish to have a reaction rule
that replaces $m$ with $m + n$. The only way to do that is by having a separate
reaction rule for each possible value of $n$. Another example of a simple
concept with no bigraphical representation is checking whether an element is in
a set, where both the element and the set have some representations in a
bigraph.

Third, there are significant limitations with regards to PRISM integration.
While we implemented the functionality to export transition rewards based on how
PRISM exports them, to the author's surprise, PRISM does not support importing
them (yet). In addition, while it is trivial to support both positive and
negative numbers as rewards/costs, for PRISM it is a completely different
problem---one that is not solved yet \cite{DBLP:conf/tacas/0001BCDDKM016}. Also,
the format for exporting the data to PRISM limits the kind of properties that
can be computed or checked: while PRISM supports having several rewards and
referring to each of them as needed, there is no way to define several rewards
within the plain text formats provided. Finally, while it would be trivial to
extend the nondeterminism to different ``players'' making different decisions
(modelling stochastic multiplayer games \cite{DBLP:phd/dnb/Ummels11}), and PRISM
has an extension that supports such scenarios \cite{DBLP:conf/tacas/ChenFKPS13},
there is no way to import player labels together with the transition system.

On the other hand, choosing to model with bigraphs rather than, e.g., the
PRISM language has some clear advantages as well. First, our example model in
Section \ref{first_agent_section} resulted in significantly less code than a
similar model in PRISM \cite{dblp:conf/nfm/giaquintahimn18}. Second, our models
can be much easier to modify: in order to change or extend the grid in some way
one only needs to change a single bigraph. Similarly, changing how the movement
of the agent is represented (the number of directions of movement, tracking
rotations separately from moving forward, adding a probability of accidentally
moving the wrong way, etc.) results in several simple changes. Third, bigraphs
come with a visual representation as well as constructs such as nesting and
linking, allowing for a richer and more intuitive way to represent many modelled
situations.

\section{Conclusions and Future Work}

\begin{figure}
  \centering
  \begin{tikzpicture}
    % outer walls
    \draw (4, 2) -- (4, 0) -- (1, 0) -- (1, 2) -- (0, 2) -- (0, 3) -- (4, 3) -- (4, 2);
    % inner walls
    \draw (2, 2) -- (3, 2) -- (3, 1);
    %\draw[red,ultra thick,dashed] (2, 2) -- (2, 1);
    % door 1
    \draw[dotted] (1.25,2) arc (180:270:0.5cm);
    \draw[brown,ultra thick] (1.75,1.5) -- (1.75,2) node[right,  near start]{};
    \draw (1, 2) -- (1.25, 2);
    \draw (1.75, 2) -- (2, 2);
    % door 2
    \draw[dotted] (3.25,2) arc (180:270:0.5cm);
    \draw[brown,ultra thick] (3.75,1.5) -- (3.75,2) node[right,  near start]{};
    \draw (3, 2) -- (3.25, 2);
    \draw (3.75, 2) -- (4, 2);
    % door 3
    \draw[dotted] (3, 0.75) arc (90:180:0.5cm);
    \draw[brown,ultra thick] (2.5, 0.25) -- (3, 0.25) node[right,  near start]{};
    \draw (3, 0.25) -- (3, 0);
    \draw (3, 0.75) -- (3, 1);
    % background grid
    \draw[gray,very thin] (1, 3) -- (1, 2);
    \draw[gray,very thin] (2, 3) -- (2, 0);
    \draw[gray,very thin] (3, 3) -- (3, 2);
    \draw[gray,very thin] (1, 1) -- (4, 1);
  \end{tikzpicture}
  \caption{An example map where both rooms have two doors.}
  \label{multiple_doors}
\end{figure}

In this paper we have proposed a way to extend BRSs with probabilities and
nondeterminism, allowing them to represent any MDP. We have presented a new
interface for the development and visualisation of BRSs and provided two
example models for autonomous agents, finding more efficient representations for
models that already exist as well as tackling new challenges.

% TODO: code for representing the map
% TODO: what if a door to a room is locked (don't remove the room itself)
Our second example in particular has introduced two new directions for modelling
space that have not been explored in full yet. First, while it is often useful
to think of one space as nested inside another, sometimes a room might have
several doors: one or two to a corridor, and one to an adjacent room (see, e.g.,
Fig. \ref{multiple_doors}). The model could support this by using the sharing
part of bigraphs with sharing\footnote{While our definitions are based on this
  extension, the additional functionality was not used.}, i.e., have the room be
inside both a \texttt{Cell} in the corridor and a \texttt{Cell} of another room.
Second, the model glues the discovery of the structure of a room together with
being near the door: once the agent visits the \texttt{Cell} encompassing the
room, it immediately discovers what is inside it. It would improve the model
significantly to have discovery as a completely separate process, where the
creator of the map has full control over when a substructure is discovered.

Finally, one ought to note that a bigraphical representation can often be
asymptotically more compact than a full MDP or discrete-time Markov chain and
sometimes even than a PRISM description of the same situation (until the BRS is
expanded into its full transition system, that is). This raises an interesting
question: can we define complexity classes of models, paying particular
attention to the MDPs that have an exponential number of states, but can be
represented in a polynomial number of bigraphs and reaction rules? Which of
these models do not have a PRISM representation of polynomial size?
